# References

<div id="refs"></div>

# (APPENDIX) Appendix {-}

# Reading and writing spatial data in R

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 100)
```

```{r, echo = FALSE}
source("package_list.R")
get.pckg.info("A01-Data-Manipulation.Rmd")
```

## Sample files for this exercise{-#app1_2}

First, you will need to download some sample files from the github repository.  Make sure to set your R session folder to the directory where you will want to save the sample files before running the following code chunks.


```{r results='hide'}
download.file("https://github.com/mgimond/Spatial/raw/main/Data/Income_schooling.zip", 
              destfile = "Income_schooling.zip" , mode='wb')
unzip("Income_schooling.zip", exdir = ".")
file.remove("Income_schooling.zip")

```


```{r}
download.file("https://github.com/mgimond/Spatial/raw/main/Data/rail_inters.gpkg", 
              destfile = "./rail_inters.gpkg", mode='wb')
```

```{r}
download.file("https://github.com/mgimond/Spatial/raw/main/Data/elev.img",  
              destfile = "./elev.img", mode='wb')               
```

## Introduction {-#app1_3}

There are several different R spatial formats to choose from. Your choice of format will largely be dictated by the package(s) and or function(s) used in your workflow. A breakdown of formats and intended use are listed below.


```{r echo=FALSE}
my_tbl <- tibble::tribble(
  ~`Data format`, ~`Used with...`, ~`Used in package...`,  ~`Used for...`, ~Comment,
  "\`sf\`", "vector",  "\`sf\`, others", "visualizing, manipulating, querying", "This is the new spatial standard in R. Will also read from spatially enabled databases such as postgresSQL.",
   "\`raster\`",  "raster", "\`raster\`, others", "visualizing, manipulating, spatial statistics", "This has been the most popular raster format fo rmany years. But, it is gradually being supplanted by `terra`",
  "\`SpatRaster\`",  "terra", "\`terra\`, others", "visualizing, manipulating, spatial statistics", "This is gradually replacing `raster`",
   "\`SpatialPoints*\` \n \`SpatialPolygons*\` \n \`SpatialLines*\` \n \`SpatialGrid*\`\n",  "vector and raster",  "\`sp\`, \`spdep\`",  "Visualizing, spatial statistics", "These are legacy formats. \`spdep\` now accepts  `sf` objects",
    "\`ppp\` \`owin\`",   "vector",   "\`spatstat\`", "Point pattern analysis/statistics", NA,
     "\`im\`",   "raster",   "\`spatstat\`", "Point pattern analysis/statistics", NA
  )


`%>%` <- magrittr::`%>%`
kableExtra::kable_styling(
        kableExtra::kable(my_tbl, digits = 3, row.names = FALSE, align = c("r", "c","c","l","l") ,
        caption = NULL, format = "html"),
        bootstrap_options = c("striped", "hover", "condensed"),
        position = "left", full_width = FALSE)  %>% 
  kableExtra::row_spec(0, color="white", background = "#6E6E6E", align="c") %>%  
  kableExtra::footnote(number=c("The `spatial*` format includes SpatialPointsDataFrame, SpatialPolygonsDataFrame, SpatialLinesDataFrame, etc...") )

```


There is an attempt at standardizing the spatial format in the R ecosystem by adopting a well established set of spatial standards known as [simple features](https://en.wikipedia.org/wiki/Simple_Features). This effort results in a recently developed package called  [`sf`](https://r-spatial.github.io/sf/) [@sf]. It is therefore recommended that you work in an `sf` framework  when possible. As of this writing, most of the _basic_ data manipulation and visualization operations can be successfully conducted using `sf` spatial objects. 

Some packages such as `spdep` and `spatstat` require specialized data object types. This tutorial will highlight some useful conversion functions for this purpose.



## Creating spatial objects {-#app1_4}

The following sections demonstrate different spatial data object creation strategies.

### Reading a shapefile {-#app1_4_1}

Shapefiles consist of many files sharing the same core filename and different suffixes (i.e. file extensions). For example, the sample shapefile used in this exercise consists of the following files:

```{r echo= FALSE}
list.files(".", "Income_schooling.*")
```

Note that the number of files associated with a shapefile can vary. `sf` only needs to be given the `*.shp` name. It will then know which other files to read into R such as projection information and attribute table.

```{r results='hide'}
library(sf)
s.sf <- st_read("Income_schooling.shp")
```

Let's view the first few records in the spatial data object. 

```{r}
head(s.sf, n=4)  # List spatial object and the first 4 attribute records
```

Note that the `sf` object stores not only the geometry but the coordinate system information and attribute data as well. These will be explored later in this exercise.

### Reading a GeoPackage {-#app1_4_2}

A geopackage can store more than one layer. To list the layers available in the geopackage, type:

```{r}
st_layers("rail_inters.gpkg")
```

In this example, we have two separate layers: `Interstate` and `Rail`. We can extract each layer separately via the `layer=` parameter.

```{r results='hide'}
inter.sf <- st_read("rail_inters.gpkg", layer="Interstate")
rail.sf  <- st_read("rail_inters.gpkg", layer="Rail")
```

### Reading a raster {-#app1_4_3}

> In earlier versions of this tutorial, the `raster` package was used to read raster files. This is being supplanted by `terra` which will be the package used in this and in subsequent exercises. 

`terra` will read many different raster file formats such as geoTiff, Imagine and HDF5 just to name a few. To see a list of supported raster file formats on your computer simply run:

```{r eval = FALSE}
terra::gdal(drivers = TRUE) |> subset(type == "raster")
```

In the following example, an _Imagine_ raster file is read into R using the `rast` function.

```{r}
library(terra)
elev.r <- rast("elev.img")
```

The object class is of type `SpatRaster`.

```{r}
class(elev.r)
```


What sets a `SpatRaster` object apart from other R data file objects is its storage. By default, data files are loaded into memory, but `SpatRaster` objects are not. This can be convenient when working with raster files too large for memory. But this comes at a performance cost. If your RAM is large enough to handle your raster file, it's best to load the entire dataset into memory.

To check if the `elev.r` object is loaded into memory, run:

```{r}
inMemory(elev.r)
```

An output of `FALSE` indicates that it is not. To force the raster into memory use `set.values`:

```{r}
set.values(elev.r)
```

Let's check that the raster is indeed loaded into memory:

```{r}
inMemory(elev.r)
```

Now let's look at the raster's properties:

```{r}
elev.r
```

The raster object returns its grid dimensions (number of rows and columns), pixel size/resolution (in the layer's coordinate system units), geographic extent, native coordinate system (UTM NAD83 Zone 19 with units of meters) and min/max raster values.

### Creating a spatial object from a data frame {-#app1_4_5}

Geographic point data locations recorded in a spreadsheet can be converted to a spatial point object. Note that it's important that you specify the coordinate system used to record the coordinate pairs since such information is not stored in a data frame. In the following example, the coordinate values are recorded in a WGS 1984 geographic coordinate system (`crs = 4326`).

```{r}
# Create a simple dataframe with lat/long values
df <- data.frame(lon = c(-68.783, -69.6458, -69.7653),
                 lat = c(44.8109, 44.5521, 44.3235),
                 Name= c("Bangor", "Waterville", "Augusta"))

# Convert the dataframe to a spatial object. Note that the
# crs= 4326 parameter assigns a WGS84 coordinate system to the 
# spatial object
p.sf <- st_as_sf(df, coords = c("lon", "lat"), crs = 4326) 
p.sf  
```

### Geocoding street addresses {-#app1_4_6}

The `tidygeocoder` package will convert street addresses to latitude/longitude coordinate pairs using a wide range of geocoding services such as the US census and Google. Some of these geocoding services will require an API key, others will not. Click [here](https://jessecambon.github.io/tidygeocoder/articles/geocoder_services.html) to see the list of geocoding services supported by `tidygeocoder` and their geocoding limitations. In the example that follows, the `osm` geocoding service is used by default.

```{r}
library(tidygeocoder)

options(pillar.sigfig = 7) # Increase significant digits in displayed output

dat <- data.frame(
  name = c("Colby College", "Bates College", "Bowdoin College"),
  address = c("4000 Mayflower drive, Waterville, ME , 04901",
              "275 College st, Lewiston, ME 04240",
              "255 Maine St, Brunswick, ME 04011"))

geocode(.tbl = dat, address = address, method = "osm")           

```

Another free (but manual) alternative, is to use the US Census Bureau's web [geocoding service](https://geocoding.geo.census.gov/geocoder/locations/addressbatch?form) for creating lat/lon values from a file of US street addresses. This needs to be completed via their web interface and the resulting data table (a CSV file) would then need to be loaded into R as a data frame.

## Converting from an `sf` object {-#app1_5}

Packages such as `spdep` (older versions only) and `spatsat` do not support `sf` objects. The following sections demonstrate methods to convert from `sf` to other formats.

### Converting an `sf` object to a `Spatial*` object (`spdep`/`sp`) {-#app1_5_1}

The following code will convert point, polyline or polygon features to a `spatial*` object. While the current version of `spdep` will now accept `sf` objects, converting to `spatial*` objects will be necessary with legacy `spdep` packages. In this example, an `sf` polygon feature is converted to a `SpatialPolygonsDataFrame` object.

```{r}
s.sp <- as_Spatial(s.sf)
class(s.sp)
```

### Converting an `sf` polygon object to an `owin` object {-#app1_5_2}

The `spatstat` package is used to analyze point patterns however, in most cases, the study extent needs to be explicitly defined by a polygon object. The polygon should be of class `owin`. 

```{r}
library(spatstat)
s.owin <- as.owin(s.sf)
class(s.owin)
```

Note the loading of the package `spatstat`. This is required to access the `as.owin.sf` method for `sf`.
Note too that the attribute table gets stripped from the polygon data. This is usually fine given that the only reason for converting a polygon to an `owin` format is for delineating the study boundary.

### Converting an `sf` point object to a `ppp` object {-#app1_5_3}

The `spatstat` package is currently designed to work with projected (planar) coordinate system. If you attempt to convert a point object that is in a geographic coordinate system, you will get the following error message:

```{r error=TRUE}
p.ppp <- as.ppp(p.sf)
```

The error message reminds us that a geographic coordinate system (i.e. one that uses angular measurements such as latitude/longitude) cannot be used with this package. If you encounter this error, you will need to project the point object to a projected coordinate system.

In this example, we'll project the `p.sf` object to a UTM coordinate system (`epsg=32619`). Coordinate systems in R are treated in a later appendix. 

```{r}
p.sf.utm <- st_transform(p.sf, 32619) # project from geographic to UTM
p.ppp <- as.ppp(p.sf.utm)             # Create ppp object
class(p.ppp)
```

Note that if the point layer has an attribute table, its attributes will be converted to `ppp` _marks_. These attribute values can be accessed via `marks(p.ppp)`.

### Converting a `SpatRaster` object to an `im` object {-#app1_5_4}

To create a `spatstat` `im` raster object from a `SpatRaster` object, you will need to first create a three column dataframe from the `SpatRaster` objects with the first two columns defining the X and Y coordinate values of each cell, and the third column defining the cell values

```{r}
df <- as.data.frame(elev.r,xy=TRUE)
elev.im <- as.im(df)
class(elev.im)
```

## Converting to an `sf` object {-#app1_6}

All aforementioned spatial formats, except `owin`, can be coerced to an `sf` object via the `st_as_sf` function. for example:

```{r results='hide'}
st_as_sf(p.ppp)  # For converting a ppp object to an sf object
st_as_sf(s.sp)   # For converting a Spatial* object to an sf object
```

## Dissecting the `sf` file object {-#app1_7}

```{r}
head(s.sf,3)
```

The first line of output gives us the geometry type, `MULTIPOLYGON`,  a multi-polygon data type. This is also referred to as a multipart polygon. A single-part `sf` polygon object will adopt the `POLYGON` geometry.

The next few lines of output give us the layer's bounding extent in the layer's native coordinate system units. You can extract the extent via the `st_bbox()` function as in `st_bbox(s.sf)`.

The following code chunk can be used to extract addition coordinate information from the data. 

```{r eval = FALSE}
st_crs(s.sf)
```

Depending on the version of the `PROJ` library used by `sf`, you can get two different outputs. If your version of `sf` is built with a version of `PROJ` older than `6.0`, the output will consist of an **epsg** code (when available) and a **proj4** string as follows:

```{r eval = FALSE}
Coordinate Reference System:
  EPSG: 26919 
  proj4string: "+proj=utm +zone=19 +datum=NAD83 +units=m +no_defs"
```


If your version of `sf` is built with a version of `PROJ` `6.0` or greater, the output will consist of a user defined CS definition (e.g. an **epsg** code), if  available, and a *Well Known Text* (WKT) formatted coordinate definition that consists of a series of `[ ]` tags as follows:

```{r echo = FALSE}
st_crs(s.sf)
```

The WKT format will usually start with a `PROJCRS[...]` tag for a projected coordinate system, or a `GEOGCRS[...]` tag for a geographic coordinate system. More information on coordinate systems in R can be found in  [the coordinate systems appendix](https://mgimond.github.io/Spatial/coordinate-systems-in-r.html).

What remains of the `sf` summary output is the first few records of the attribute table. You can extract the object's table to a dedicated data frame via:

```{r}
s.df <- data.frame(s.sf)
class(s.df)
head(s.df, 5)
```

The above chunk will also create a geometry column. This column is somewhat unique in that it stores its contents as a **list** of geometry coordinate pairs (polygon vertex coordinate values in this example).

```{r}
str(s.df)
```

You can also opt to remove this column prior to creating the dataframe as follows:

```{r}
s.nogeom.df <- st_set_geometry(s.sf, NULL) 
class(s.nogeom.df)
head(s.nogeom.df, 5)
```

## Exporting to different data file formats {-#app1_8}

You can export an `sf` object to many different spatial file formats such as a shapefile or a geopackage.

```{r eval=FALSE}
st_write(s.sf, "shapefile_out.shp", driver = "ESRI Shapefile")  # create to a shapefile 
st_write(s.sf, "s.gpkg", driver = "GPKG")  # Create a geopackage file
```

If the file you are writing to already exists, the above will throw an error. To force an overwrite, simply add the `delete_layer = TRUE` argument to the `st_write` function.

You can see a list of writable vector formats via:

```{r eval = FALSE}
gdal(drivers = TRUE) |> subset(can %in% c("write", "read/write" ) & type == "vector")
```

The value in the `name` column is the driver name to pass to the `driver =` argument in the `st_write()` function.

To export a raster to a data file, use `writeRaster()` function.

```{r eval=FALSE}
writeRaster(elev.r, "elev_out.tif", gdal = "GTiff" ) # Create a geoTiff file
writeRaster(elev.r, "elev_out.img", gdal = "HFA" )  # Create an Imagine raster file
```

You can see a list of writable raster formats via:

```{r eval = FALSE}
gdal(drivers = TRUE) |> subset(can %in% c("write", "read/write" ) & type == "raster")
```

The value in the `name` column is the driver name to pass to the `gdal =` argument in the `writeRaster()` function.

```{r remove_temp_files, echo = FALSE, results='hide', message=FALSE}
files.rm <- dir(".", "Income_schooling.*") 
file.remove(file.path(".", files.rm ))

file.remove(file.path(".", "elev.img" ))

file.remove(file.path(".", "rail_inters.gpkg" ))
```


